---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

While the categories do say what the crimes are, they are still pretty broad. This can make it difficult to understand the data and examine the details of the crime. For example, there are two categories detailing the number of thefts in the county, however only one specifies that it was a theft from a car, versus thefts in other places. However, I think the theft and stolen auto categories are definitely the most important from a reporting stand point because they occur at significantly higher rates than other crimes. 

### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_totals <- pg_crime |> mutate(total_incidents = (`Jan`+`Feb`+`Mar`+`Apr`+`May`+`Jun`+`Jul`+`Aug`+`Sep`+`Oct`+`Nov`+`Dec`))
pg_crime_averages <- pg_crime_totals |> mutate(average_incidents = total_incidents / 12)
```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

The category that has the highest number of incidents per month is titled "accidents." I think this makes sense because it is not a crime to get into a car accident and they are very common. However, the crime with the greatest number of incidents is auto theft, while the category with the fewest number of incidents was residential robbery. I think this shows that there are a high number of incidents involving cars in Prince George's County, likely because it is easier for people to have access too. However, it is interesting that while there was an average of about 400 car thefts a month, only about 40 were recovered.  


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |>
  mutate(june_change   = (`Jun` - `average_incidents`) / `average_incidents` * 100,
    july_change   = (`Jul` - `average_incidents`) / `average_incidents` * 100,
    august_change = (`Aug` - `average_incidents`) / `average_incidents` * 100)

```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

While most of the categories had some increases in the number of incidents during the summer months, the only category that had a consistent increase throughout each summer month was the homicide category. Because there were not consistent increases during every month and many of the categories saw a decrease in the number of crimes, I think this data disproves the theory that crime is more likely to occur in the summer months. However, it does show that the number of homicides is greater during the summer months. 

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Prince George's County sees an uptick in homicides during the summer months. 

Prince George's County Police Department crime data shows there were 8 incidents in June and 7 in July and August. This is an increase from the monthly average of about 5.8 incidents. 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
