---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2025.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

The data sheet provided the name of the employee, their employee status, salary, how often they are paid and their title. Most of the salary values seemed consistent for what I would expect a White House employee would be paid. However, I did see that the salaries for eight of the employees were zero. This definitely stuck out to me because it is not likely they are working for free and many of those employees were high profile members of Trump's administration. 

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |>
  summarise( 
  mean_wh_salary = mean(SALARY),
  median_wh_salary = median(SALARY))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |>
  summarise( 
  mean_wh_salary = mean(SALARY),
  median_wh_salary = median(SALARY), 
  min_wh_salary = min(SALARY),
  max_wh_salary = max(SALARY))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

Based on my analysis, the mean and median salaries show that a significant number of White House employees make about $100,000 a year. However, I noticed that only one employee made anywhere close to the maximum salary of $225,700. I think the most important aspect of this data is that some White House employees are recorded as not being given a salary. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

Based on the data, I think a good story idea to pursue would be to see why there were some White House employees who were recorded as not making a salary and how much they make from their work for the White House. The next step I would take to develop this idea is to see if they are working for other federal departments and send in a data request to see how much they are making there. 
