---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```
### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
mandms <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```
### Task 3: Sample 5 rows of data from the dataset
```{r}
set.seed(42)

sample_5 <- mandms |> sample_n(5)

proportion_5 <- mean(sample_5$red)
```

### Task 4: Calculate the mean age and standard deviation of the sample, using either the blue or white number of M&Ms

```{r}
sample_5 |> summarise(mean_red = mean(red), sd_red = sd(red))

```

### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4
```{r}
mandms  |> summarise (mean_red = mean(red), sd_red = sd(red))
```
### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?

The sample mean is greater than the population mean. The sample mean is 79.6, while the population mean is 72.98. Additionally, the sample standard deviation is 10.06 and the population's is 16.46. This means that the data in the sample group is clustered closer together than the population data. 

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

The data for the sample group showed that the average number of red M&Ms is greater than the average for the population. This shows that even though collecting data through random sampling can be an indicator for what a population looks like, it is not a perfect representation. As the sample becomes larger, the better it represents the population. Additionally, because the sample group only had 5 rows of data, it makes sense that the sample standard deviation was lower than the population's standard deviation. This means that there was lower variability in the data because it was a smaller sample. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
