---
title: "final_project"
output: html_document
---

### load the data 
```{r}
library(tidyverse)

spotify <- read_csv("spotify-2023.csv")
```

The data set shows a list of the Spotify's 953 most famous songs throughout 2023. For my analysis, I want to see if a song's danceability has any effect on its popularity. The data includes several variables that could be used in my analysis, including the chart ranking and the number of beats per minute the song has. I am mainly going to use the dataset's danceability column and the playlist count column in my analysis. 

Based on the feedback I received from the last assignment, I am planning to conduct a multiple regression analysis to see what other factors could impact a song's popularity (Spotify playlist count). 

For the final project, how many different types of analysis should I conduct?

### Sunday, April 19 - Tests to Run 
For this project I think I would start by making a scatterplot showing the relationship between the danceability and the playlist count. I would also add a trend line to the graph. This would help me see what type of relationship these two variables have. 

I would also run a linear regression analysis between these two factors to find the p-value, which would help me see if danceability has a statistically significant effect on popularity. I would also like to see if the effect is meaningfully significant. 

I am also planning to run a multiple regression using the beats per minute and the chart ranking to see if that also has an impact on popularity (playlist count). 

While I think these tests will be helpful, I would like to confirm that these would be good to use in my analysis.

### Monday, April 27 - 

#load the tidyverse 
```{r}
library(tidyverse)
```

#Mean and standard deviation of playlist count  
```{r}
spotify |> summarize(mean = mean(in_spotify_playlists), sd = sd(in_spotify_playlists))
```

#Make a histogram of the playlist count 
```{r}
spotify |> 
  ggplot() + 
  geom_histogram(aes(x = in_spotify_playlists), binwidth = 1000)
```

#Add a line to the histogram for the mean and standard deviation playlist counts
```{r}
spotify|> 
  ggplot() + 
  geom_histogram(aes(x = in_spotify_playlists), binwidth = 1000) +
  geom_vline(aes(xintercept = mean(in_spotify_playlists)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(in_spotify_playlists) - sd(in_spotify_playlists)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(in_spotify_playlists) + sd(in_spotify_playlists)), color = "blue", linetype = "dashed", size = 1)
```

The histogram is rightly skewed, meaning that fewer Spotify songs are added to more than about 15,000 playlists. The histogram shows that songs in this data set are added to an average of about 5200 playlists total. 

#Linear Regression 
```{r}
log(spotify$in_spotify_playlists)
model <- lm(log(spotify$in_spotify_playlists) ~ spotify$`danceability_%`, data = spotify)
summary(model)
```

 
```{r}
spotify |>
  ggplot(aes(x = `danceability_%`, y = log(in_spotify_playlists))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Danceability (%)", y = "Playlist Counts",
       title = "Relationship Between Danceability Score and Playlist Count")
```
As the p-value is 0.0005, which is less than the significance value, the linear regression shows that there is a statistically significant relationship between a song's danceability score and the number of playlists it is added to. It looks like there is a slightly negative relationship between danceability and playlist count. However, the plots look like they are scattered around. 

